Online Multiclass Linear Classification with Bandit Feedback

### 2018-11-15 arxiv-version/main.tex
Lower bound for multiclass perceptron

In this paper, we focus on the special case when the examples are \emph{linearly
separable with a margin}. In the full-information feedback setting, it is well
know that if the examples have norm at most $R$ and are linearly separable with
a margin $\gamma$ then the \textsc{Multiclass Perceptron} algorithm makes at
most $\lfloor 2(R/\gamma)^2 \rfloor$ mistakes. This result is very satisfactory
since the upper bound on the number of mistakes is information-theoretically
optimal and at the same time the \textsc{Multiclass Perceptron} algorithm has
low time and memory complexity.

The bandit feedback setting, however, is much more challenging. We start by
dissecting the notion of linear separability. We define two notions of linear
separability for multiclass classification: the standard linear separability
and \emph{strong linear separability}. For binary classification, these two
definitions are identical. However, they differ for multiclass classification.
For multiclass classification with $K$ classes the standard notion of linear
separability means that the examples from each class lie in an intersection of
$K-1$ halfspaces and the examples outside the class lie in the complement of the
intersection of the halfspaces. Strong linear separability means that examples
from each class can be separated from the remaining examples by a \emph{single}
hyperplane. We state the definitions precisely
in Section~\ref{section:notions-of-linear-separability}.

If the examples are strongly linearly separable, there is a simple and efficient
algorithm for the bandit feedback setting with expected number of mistakes
at most $K (R/\gamma)^2$. The algorithm can be viewed as running $K$ copies of
the \textsc{Binary Perceptron} algorithm, one copy for each class.

For the standard notion of linear separability, we propose three algorithms with
various time-complexity-versus-mistake-bound trade-offs. Two of these algorithms
map the examples into a high-dimensional vector space induced by a kernel. In
this space the examples become \emph{strongly} linearly separable. This allows
us to run a kernelized version of the algorithm for the strongly linearly
separable case.

@@

The notion of linear separability with a margin is standard. It used in the
full-information setting to upper bound the number of mistakes of the
\textsc{Multiclass Perceptron} algorithm. For completeness, we state the
algorithm below. It is a folklore result that if a set of labeled examples is
separable with a margin $\gamma$ and the norm of the examples is bounded by $R$
then \textsc{Multiclass Perceptron} algorithm makes at most $\left\lfloor
(R/\gamma)^2 \right \rfloor$ mistakes. Another folklore result is that
\textsc{Multiclass Perceptron} is an optimal algorithm in the sense that any
deterministic algorithm makes $\left\lfloor (R/\gamma)^2 \right \rfloor$
mistakes in the worst case. For completeness, we give proofs  in
Section~\ref{section:multiclass-perceptron-proofs} for both of these results.

\begin{algorithm}[h]
\caption{\textsc{Multiclass Perceptron}
\label{algorithm:mutliclass-perceptron}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$.}
\STATE{Initialize $w_1^{(1)} = w_2^{(1)} = \dots = w_K^{(1)} = 0$}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t$}
\STATE{Predict $\widehat y_t = \argmax_{i \in \{1,2,\dots,K\}} \ip{w_t^{(i)}}{x_t}$}
\STATE{Observe $y_t \in \{1,2,\dots,K\}$}
\IF{$\widehat y_t \neq y_t$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\}$}
\STATE{Update $w_{y_t}^{(t+1)} = w_{y_t}^{(t)} + x_t$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} - x_t$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[h]
\caption{Algorithm for strongly linearly separable examples
\label{algorithm:algorithm-for-strongly-linearly-separable-examples}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$.}
\STATE{Initialize $w_1^{(1)} = w_2^{(1)} = \dots = w_K^{(1)} = 0$}
\FOR{$t=1,2,\dots,T$}
\STATE{Receive feature vector $x_t$}
\STATE{Compute $S_t = \left\{ i ~:~ 1 \le i \le K, \ \ip{w_i^{(t)}}{x_t} \ge 0 \right\}$}
\IF{$S_t = \emptyset$}
\STATE{Predict $\widehat y_t \sim \text{Uniform}(\{1,2,\dots,K\})$}
\ELSE
\STATE{Predict $\widehat y_t \sim \text{Uniform}(S_t)$}
\ENDIF
\STATE{Receive feedback $z_t \in \{0,1\}$ where $z_t = \indicator{\widehat y_t \neq y_t}$}
\IF{$S_t = \emptyset$}
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} + x_t$}
\ENDIF
\ELSE
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} - x_t$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ENDIF
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

@@

\begin{theorem}[Upper bound]
Let $(V, \ip{\cdot}{\cdot})$ be an inner product space, let $K$ be a positive
integer, let $\gamma$ be a positive real number and let $R$ be a non-negative real
number. If $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is a sequence of labeled
examples in $V \times \{1,2,\dots,K\}$ that are linearly separable with margin
$\gamma$ and $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
then \textsc{Multiclass Perceptron} algorithm makes at most $\lfloor
2(R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

@@

\begin{theorem}[Lower bound]
Let $K$ be a positive integer, let $\gamma$ be a positive real number and let
$R$ be a non-negative real number. For any deterministic algorithm for the
full-information multiclass classification there exists an inner product space
$(V, \ip{\cdot}{\cdot})$, a non-negative integer $T$ and a sequence of labeled
examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ examples in $V \times
\{1,2,\dots,K\}$ that are linearly separable with margin $\gamma$, the norms
satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$ and the algorithm
makes at least $\lfloor 2(R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

>>>>

In this paper, we focus on the special case when the examples lie in $\R^d$ and
are \emph{linearly separable with a margin}. In the full-information feedback
setting, it is well know that if the examples have norm at most $R$ and are
linearly separable with a margin $\gamma$ then the \textsc{Multiclass
Perceptron} algorithm makes at most $\lfloor 2(R/\gamma)^2 \rfloor$ mistakes.
This result is very satisfactory since the upper bound on the number of mistakes
is information-theoretically optimal and at the same time the \textsc{Multiclass
Perceptron} algorithm has low time and memory complexity.

The bandit feedback setting, however, is much more challenging. We start by
dissecting the notion of linear separability in the multiclass setting. Beside
the standard notion of linear separability, we define also and \emph{strong
linear separability}. For binary classification, these two definitions are
identical. However, they differ for multiclass classification. For multiclass
classification with $K$ classes the standard notion of linear separability means
that the examples from each class lie in an intersection of $K-1$ halfspaces and
the examples outside of the class lie in the complement of the intersection of
the halfspaces. Strong linear separability means that examples from each class
can be separated from the remaining examples by a \emph{single} hyperplane. We
state the definitions precisely in
Section~\ref{section:notions-of-linear-separability}.

For the case when the examples are strongly linearly separable, we design a
simple and efficient algorithm with expected number of mistakes at most $K
(R/\gamma)^2$. The algorithm can be viewed as running $K$ copies of the
\textsc{Binary Perceptron} algorithm, one copy for each class. Intuitively
speaking, the factor $K$ is the price we pay for the bandit feedback,
or more precisely, the lack of full-information feedback.

For the standard notion of linear separability, we propose three algorithms with
various time-complexity-versus-mistake-bound trade-offs. Two of these algorithms
map the examples into a high-dimensional inner product space induced by a
kernel. In this space the examples become \emph{strongly} linearly separable.
This allows us to run a kernelized version of the algorithm for the strongly
linearly separable case.

@@

The notion of linear separability with a margin is standard. It used in the
full-information setting to upper bound the number of mistakes of the
\textsc{Multiclass Perceptron} algorithm. For completeness, we state the
algorithm below. It is a folklore result that if a set of labeled examples is
separable with a margin $\gamma$ and the norm of the examples is bounded by $R$
then \textsc{Multiclass Perceptron} algorithm makes at most $\left\lfloor
2(R/\gamma)^2 \right \rfloor$ mistakes. Another folklore result is that
\textsc{Multiclass Perceptron} is an optimal algorithm in the sense that any
deterministic algorithm makes $\left\lfloor 2(R/\gamma)^2 \right \rfloor$
mistakes in the worst case. For completeness, we give proofs  in
Section~\ref{section:multiclass-perceptron-proofs} for both of these results.

\begin{algorithm}[h]
\caption{\textsc{Multiclass Perceptron}
\label{algorithm:mutliclass-perceptron}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$. Inner product space $(V,\ip{\cdot}{\cdot})$.}
\STATE{Initialize $w_1^{(1)} = w_2^{(1)} = \dots = w_K^{(1)} = 0$}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t \in V$}
\STATE{Predict $\widehat y_t = \argmax_{i \in \{1,2,\dots,K\}} \ip{w_t^{(i)}}{x_t}$}
\STATE{Observe $y_t \in \{1,2,\dots,K\}$}
\IF{$\widehat y_t \neq y_t$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{y_t, \widehat y_t\}$}
\STATE{Update $w_{y_t}^{(t+1)} = w_{y_t}^{(t)} + x_t$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} - x_t$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[h]
\caption{\textsc{Bandit Algorithm for Strongly Linearly Separable Examples}
\label{algorithm:algorithm-for-strongly-linearly-separable-examples}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$. Inner product space $(V,\ip{\cdot}{\cdot})$.}
\STATE{Initialize $w_1^{(1)} = w_2^{(1)} = \dots = w_K^{(1)} = 0$}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t \in V$}
\STATE{Compute $S_t = \left\{ i ~:~ 1 \le i \le K, \ \ip{w_i^{(t)}}{x_t} \ge 0 \right\}$}
\IF{$S_t = \emptyset$}
\STATE{Predict $\widehat y_t \sim \text{Uniform}(\{1,2,\dots,K\})$}
\ELSE
\STATE{Predict $\widehat y_t \sim \text{Uniform}(S_t)$}
\ENDIF
\STATE{Observe feedback $z_t \in \{0,1\}$ where $z_t = \indicator{\widehat y_t \neq y_t}$}
\IF{$S_t = \emptyset$}
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} + x_t$}
\ENDIF
\ELSE
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} - x_t$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ENDIF
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

@@

\begin{theorem}[Mistake upper bound]
Let $(V, \ip{\cdot}{\cdot})$ be an inner product space, let $K$ be a positive
integer, let $\gamma$ be a positive real number and let $R$ be a non-negative real
number. If $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is a sequence of labeled
examples in $V \times \{1,2,\dots,K\}$ that are linearly separable with margin
$\gamma$ and $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
then \textsc{Multiclass Perceptron} algorithm makes at most $\lfloor
2(R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

@@

\begin{algorithm}[h]
\caption{\textsc{Full-Information Multiclass Classification}
\label{algorithm:mutliclass-classification}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$.}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t$}
\STATE{Predict $\widehat y_t \in \{1,2,\dots,K\}$}
\STATE{Observe $y_t \in \{1,2,\dots,K\}$}
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

\begin{theorem}[Mistake lower bound]
Let $K$ be a positive integer, let $\gamma$ be a positive real number and let
$R$ be a non-negative real number. For any deterministic algorithm for the
\textsc{Full-Information Multiclass Classification} problem there exists an
inner product space $(V, \ip{\cdot}{\cdot})$, a non-negative integer $T$ and a
sequence of labeled examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$
examples in $V \times \{1,2,\dots,K\}$ that are linearly separable with margin
$\gamma$, the norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T} \le R$
and the algorithm makes at least $\lfloor (R/\gamma)^2 \rfloor$ mistakes.
\end{theorem}

\begin{proof}
Let $T = \lfloor (R/\gamma)^2 \rfloor$ and let $x_1, x_2, \dots, x_T$ be the
orthogonal vectors such that $\norm{x_t} = R$ for all $t=1,2,\dots,T$. For
example, we can take $V = \R^T$ and $x_t = R e_t$ where $e_t$ is $t$-th element
of the standard orthonormal basis of $\R^T$.

Since the algorithm is deterministic, we can construct the sequence of labels
$y_1, y_2, \dots, y_T$ adaptively based on the predictions $\widehat y_1,
\widehat y_2, \dots, \widehat y_T$ of the algorithm. We define $y_t$ to be any
element of $\{1,2,\dots,K\}$ not equal to $\widehat y_t$. This way the algorithm
makes a mistake in every round $t=1,2,\dots,T$.

It remains to show that the examples we have constructed are linearly separable
with margin $\gamma$. To prove that we demostrate vectors $w_1, w_2, \dots, w_K$
satisfying conditions \eqref{equation:linear-separability-1} and
\eqref{equation:linear-separability-2}. We define
$$
w_i = \frac{1}{R\sqrt{T}} \sum_{\substack{t : 1 \le t \le T \\ y_t = i}} x_t \qquad \qquad \text{for $i=1,2,\dots,K$.}
$$
Let $a_i = |\{ t ~:~ 1 \le t \le T, \ y_t = i \}|$ be the number of occurences of label $i$.
It is easy to see that
$$
\norm{w_i}^2 = \frac{1}{R^2 T} \sum_{\substack{t : 1 \le t \le T \\ y_t = i}} \norm{x_t}^2 = \frac{a_i}{T} \qquad \qquad \text{for $i=1,2,\dots,K$.}
$$
Since $\sum_{i=1}^K a_i = T$, the condition
\eqref{equation:linear-separability-1} holds. To verify condition
\eqref{equation:linear-separability-2} consider any labeled example $(x_t,
y_t)$. Then, by definition of $w_{y_t}$ we have
$$
\ip{w_{y_t}}{x_t}
= \frac{1}{R\sqrt{T}} \sum_{\substack{s : 1 \le s \le T \\ y_s = y_t}} \ip{x_s}{x_t}
= \frac{1}{R\sqrt{T}} \norm{x_t}^2
= \frac{R}{\sqrt{T}}
\ge \frac{R}{R/\gamma}
= \gamma \; .
$$
and for any $i \in \{1,2,\dots,K\} \setminus \{y_t\}$ we have
$\ip{w_i}{x_t} = 0$. Therefore, condition \eqref{equation:linear-separability-2} holds.
\end{proof}
